Statistical Grand Slams

### 2023-12-14 Lahman Analysis/Formal Analysis/Draft1.qmd
Revised through beta

If modeling for the sake of learning was not reason enough to embark on this endeavor, there is the possibility of financial compensation also looming with this research question. If one is able to accurately predict the HR's that a player hits in a season, then they would probably be able to gain employment for a MLB team. HR's are the pinnacle of a hitter's trip to the plate, they are guaranteed to score at least one run and can lead to up to four being scored depending on how many people are on base. Thus, MLB teams which seek to maximize wins are likely looking to hire analysts who can accurately predict the HR's a player would hit so that they know who to target for drafting, signing, and trading. In addition, sports betting has recently become legal in many places in the United States. Thus, if someone is able to

@@

In Major League Baseball (MLB) the team on offense is allowed to send one player, at a time, up to home plate to "bat." Any time a player is sent up to bat this constitutes a plate appearance (PA). Depending on the outcome of the PA it may or may not be considered an at-bat (AB). If the PA ends in the player being walked (BB \[Base on Balls\]) then it is not recorded as an AB because it is thought the player did not receive a fair chance at hitting the baseball. On the other hand, if the PA ends in the player getting a hit or getting out, then it is assumed that he had a fair chance at hitting the ball and the PA is recorded as an AB. This distinction to the average fan is mostly arbitrary. However, in a modeling context it is very important. When modeling a player's ability to hit home runs (HR), as this paper is concerned with, the distinction is important because it is unfair to consider all the PA's of a player because this set of outcomes includes outcomes where it is assumed that the player has not been given a fair chance to hit the ball. For similar reasons, considering the subset of outcomes where it is assumed that they player has been given a fair chance to hit the ball seems more appropriate, this subset is AB's. Thus, this paper uses AB's as the main predictor for home runs hit by a certain player. Fellingham and Fisher (2017) also agree with this reasoning and use AB's, rather than PA's, in their research.

Furthermore, it is logical to ask: why not consider both? Incorporating more predictors can make a model more accurate at prediction, which both Fellingham and Fisher (2017) and this paper are concerned with. The reasons this paper does not consider both is because, especially in a Bayesian context, researchers must be concerned with multicollinearity. Without a doubt there is a strong correlation between the two because one is a subset of the other. In fact, we can see that the correlation between the two is quite strong in @fig-BattingCor. When multicollinearity is introduced into a model, it can be difficult for the Markov Chain Monte Carlo (MCMC) algorithm to parse out which of the two variables is responsible for the changes in the dependent variable. Thus, we neglect PA's from our model both because theory stipulates it is an inferior predictor and used in tandem with our current predictor it would cause problems for the modeling software.

An interesting quirk of baseball is that, like soccer, stadiums differ in dimensions. This is not universal; baseball fields and stadiums must have the same measurements for their infields. Yet, for the most part baseball stadiums are allowed to configure their outfields as they please. This becomes important because most HR's are achieved by hitting the baseball over the outfield fence. Thus, fences closer to home plate generally lead to more HR's for all hitters, ceteris paribus. The non-standardized nature of stadiums' fences gives them a variety of unique shapes and ranging outfield sizes. For example, Fenway Park (1923-) is known to have the closest outfield wall with their "Green Monster" in left field being only 305 ft. from home plate. While the Polo Grounds (1891-1963) boasted one of the furthest walls from home plate at 483 ft. in center field. Stadiums also differ in location. Marlins Park (2012-2020) sits at 15 ft. above sea level while Coors Field (1995-) sits at 5183 ft. above sea level. As illustrated there are not only vastly different dimensions that baseball fields can take on but also different environments which affect a player's ability to hit home runs. Thus, any model which attempts to model players' ability to hit HR's needs to consider ballpark factors.

MLB, as we know it, has been played since about 1871 and the origins of the game go back even father than that. Over this time the some rules of the game have undergone incredible transformations which vastly change the nature of the game. For example, from 1900-1920 baseball was in the "deadball era." During this period the materials used to create baseballs made it so that the ball absorbed much more force than it does today. This had a plethora of implications for the game and how it was played; HR's were much less common because they were harder to hit. In contrast, the period from 1994-2004 is known informally as the "steroid era" because of the rampant use of performance enhancing drugs (PED) used by players. This, in combination with other factors such as harder materials used to make baseballs, led to players obliterating a variety of previous hitting records, such as the career and season home runs records. It is obvious from these facts that decade played for player's had a drastic impact on their ability to produce outcomes. In fact, this can be visualized by considering all the runs scored during a given season from 1871-2019 displayed in @fig-RunsByYear.

Finally, age is an important factor in athletic ability. To an extent in MLB, seniority can act both as a blessing and a curse. On the one hand, older players are less able to hit home runs because their body is starting to lose its peak athletic performance it had when it was, say, 30. On the other hand, these older players also have more experience than younger players and thus may be better able to hit home runs from strategies they have found during their time playing. Further we can visualize this by looking at the median home runs hit by each player from 1871-2022, which is displayed in @fig-HR.By.Age.All. However, this graph does not depict some of the nuances of the data well. For example, we have almost no data on how often 18 -year-olds hit HR's because they play in MLB infrequently. As can be seen with the standard error bars on the ages above 40, older players suffer a similar problem of small sample size. Instead, consider @fig-HR.By.Age.20to40 which shows only considers players who are 20-40 years old during the season. This graph clearly shows the complex non-linear relationship between player performance and age. It shows that players are generally improving in their ability to hit HR's from 20-30 years old. Then, however, we see a slight downwards trend followed by an upward trend with increasing error bars. This shows that as players age past 30 their abilities start to decline. However, the jump afterwards can probably be attributed to players who are very good, continuing their careers. If a player is very good then they are going to be able to perform at above average when they are, say, 35. However, if the player is not generally considered good, then a team will likely opt to try out a new young rookie who may not be as good in the moment but is more projectable. Thus, we see less players playing past 35 which explains the higher standard errors, but they are of a higher caliber than any other age group which is why they are still playing, thus we see their proportion as higher.

@@

As mentioned, Fellingham and Fisher (2017) have both considered and published their thoughts on this research question, which concerns predicting player home run performance using Bayesian multilevel modeling. To do this, Fellingham and Fisher (2017) used data from the Lahman data set spanning 1871 to 2008 to estimate the parametric weights for their model. Over this time frame they only considered players who had played for at least 6 seasons and who, in each of those seasons, had at least 50 AB's. They further restricted their pool of players to only considered seasons where the players were 18-years-old to 45-years-old.

To predict a player's HR's they used a binomial logistic model. This model predicts HR's using the AB's of a player for the $n$ trials and the underlying ability of a player to hit a HR, $\pi$ as their probability of success. Fellingham and Fisher (2017) estimate a player's ability to hit home runs from his age in a season, decade born, season played, and home ball park. This leaves Fellingham and Fisher with the following model, which we have adapted to make model parameters the same across both their model and my model (in @sec-Model):

@@

The parameter's $\beta_{n,1-4}$ describe the weights of the orthogonal polynomial, up to the quartic, for the ages of the players. We theorize, because Fellingham and Fisher (2017) did not explicitly articulate why, they chose a quartic model because it has a closer fit to the data displayed in @fig-HR.By.Age.All and @fig-HR.By.Age.20to40. The properties of a squared term capture the rise and fall of player's HR probability over their career. The inclusion of the cubic term allows for the modeling of multiple peaks, while the quartic term allows it to maintain that overall trend up and then down seen in the data. This appears in @fig-HR.By.Age.20to40 where players see a steady increase in their HR probability during their career with a peak when they are 27. Then, their probability decreases a little, followed quickly by another increase, which then leads to another peak when the player is 31. Afterwards, the probability continues in a downward trend. This will be discussed in more depth in @sec-Model but our model, which aims to improve Fellingham and Fisher's (2017) model, does not account for the cubic or quartic effects over fear of over-fitting the age data, since the distinction between the humps is rather small.

@@

In addition, $\alpha_n$ is not the only player specific parameter of this model. The parameters $\beta_{n,1-4}$ attempt to measure how a player's age might affect their ability to hit HR's. In contrast to $\alpha_n$, this parameter does not use multilevel modeling techniques. Instead, it relies on a more traditional statistical approach which considers how age affects players aggregately. That is, it relies on one fit for all the age data and then applying that uniformly to all players considered who are that age.

In contrast, the parameter $\xi_i$ (which is $\delta_i$ in the actual paper) represents the effect that season $i$ has on player $n$'s ability to hit a HR at stadium $p$. This $\xi$ is calculated for each season considered where $i\in\{1,…,138\}$ for the 138 seasons considered. This $\xi_i$ is calculated independent of each player so it represents the different changes that we see in HR ability over all years considered. An example of why this effect is important refer to @fig-HRByYear, which clearly shows how HR rates have differed drastically over the years considered.

The parameter $\delta_p$ (which is $\xi_p$ in the actual paper) is the effect that playing in park $p$ has on the HR probability $\pi_{nip}$ in year $i$. Fellingham and Fisher (2017) decided that they want this estimate to independent of the hitter. Hence, they use no pooling or multilevel effects in this part of the model. They estimate the effect that each park has on a player's ability to hit a HR and then add the associated effect to the players who play at that ballpark in a given year.

One of the main reasons that Fellingham and Fisher's (2017) model is so strange are the priors that they chose. Disregarding that there is no justification for these priors in their paper, Fellingham and Fisher (2017) set the following priors for these parameters:

@@

These priors are quite strange because they are so uninformative. For example, consider the priors on either $\xi_i$ or $\delta_p$. The priors state that on a logistic scale, it is normal for the decade of birth and season of play parameters to vary along the interval \[-6.5, 1.5\]. On the normal probability scale, this corresponds to probability fluctuations of about (0.0015, 0.81). Furthermore, if you only consider the means that Fellingham and Fisher have used here, then you would be left with a player who, on the logit scale, hits HR's at a probability of -7.5. On the normal probability scale this becomes about 0.00055, which is 0.055% chance any AB ends in a HR. Confusingly, this is not even close to the 2.8% that Fellingham and Fisher (2017) even noted in their paper. Surely there must be better priors than this which Fellingham and Fisher (2017) could have used since it does not seem that even they believe these priors that they have set. Coincidentally, this is probably why in using MCMC to estimate their parameters they needed to use 200,000 iterations. Likely, their model could have estimated their parameters much more easily had they given it stronger priors. This is one of the areas this paper seeks to improve Fellingham and Fisher's (2017) methods.

Another questionable choice that Fellingham and Fisher (2017) make is that they estimated their model parameters using FORTRAN. Given the antiquity of FORTRAN, replication of Fellingham and Fisher (2017) is quite difficult. Not to mention that the 200,000 iterations makes the research so computationally expensive that it makes it nearly impossible for any other research team to replicate. This is disappointing because the scientific method relies on replication for the advancement of knowledge. Thus, our research seeks to make the algorithms used in this research more accessible because we chose to use R to interface with the modeling software Stan, which is written in the (more contemporary) C++ programming language.

@@

Given all the previous problems with Fellingham and Fisher (2017) mentioned above, there is an argument to be made that a simple generalized linear model (GLM) could be used instead of their complicated Bayesian semiparametric model. While it is true that a GLM could be used to predict HR's of a player it would not nearly be as accurate as a Bayesian model. Nor would it be able to as accurately capture the non-linearities present in the data. This is because the greatest strength of using the Bayesian technique is that it allows for multilevel models to be created. That is, Bayesian models allow for models to "learn" from other players and group certain players when there is not as much data available for them. In turn, this leads to better predictions about players where data is lacking. This "learning" is known as partial pooling, where the model identifies that parameters for each player (or year etc.) probably comes from the same distribution but may not share the exact same estimates.

@@

In this model, we say that the $HR$'s of player $n$ follow a binomial distribution with number of trials for player $n$ in season $i$, where player $n$ played in park $p$, are $AB_{nip}$. The probability of success (hitting a HR) is denoted as $\pi_{nip}$ for player $n$ in season $i$ playing at park $p$. Furthermore, we measure $\pi$ on the logistic scale.

As discussed in @sec-BaseBack, there are many factors that influence player $n$'s underlying probability of hitting a $HR$ in a given $AB$. The first parameter which affects this probability, $\pi_{nip}$, is $\alpha_n$. In our model $\alpha_n$ can be generally thought of as the term which is used to account for a player's innate ability to hit HR's. From our exploratory observations of the data, we noticed that the average (median or mean) proportion of AB's which resulted in HR's to be about 0.03. This means that about 3 in every 100 AB's result in HR's. This can be seen in @fig-HR.By.Age.20to40, @fig-pDist, and @fig-pDistSmall. On the logit scale, which is what our model relies on, a probability of 0.03 is about -3.5. We recognize that some players, for whatever reason, have better "innate" abilities to hit HR's than others. Therefore, we used a multilevel modeling technique for $\alpha_n$ to capture that player-to-player difference. Instead of simply using the average probability as our intercept, which is what a dummy variable would do in a GLM, we allow our model choose an intercept which adapts depending on the player and the data. That is, we give our model priors that we would expect to see in the data and then, as the model interprets additional observations, it updates these priors using Bayes theorem. For these priors, we let $\alpha$ be normally distributed with mean $\mu_0$ (where $\mu_0$ is normally distributed with mean -3.5 and standard deviation 0.1) and standard deviation $\sigma_0$ (where the parameter $\sigma_0$ itself is exponentially distributed). This allows us to have separate intercepts for each player $n$, but these intercepts come from the same hyper-priors $\mu_0$ and $\sigma_0$ that the model considers for all other players. Thus, the model constructs its posterior predictions based on both the priors and the data. In effect, this makes the model appear as though it is "learning" from the data because it has the same general sense of how a player should perform, but then adapts that prediction based on the data it takes into account to form its posterior. Specifically, for $\alpha_n$ we say that the means of all players roughly follow the distribution of $\mu_0$, but we also tell the model that it should look at how player $n$ performed to gauge its posterior prediction. A similar process is followed for $\alpha_n$'s standard deviation, $\sigma_0$.

The next parameter we considered which affects $\pi_{nip}$ was $\beta_n$ which represents the effect centered age (30 minus the player's age) had on player $n$'s ability to hit HR's. This topic was briefly touched upon in @sec-BaseBack, where we discussed how a player's age can impact their their strategy and physical prowess. These ideas are well demonstrated in @fig-HR.By.Age.20to40. In @fig-HR.By.Age.20to40, one can easily see that, in general, players' ability to hit HR's improves over the course of their career and peaks when players reach about 30 years old. After this peak, a player's ability starts to decline for the rest of their career. We chose to center the the age of the all players at 30 for several reasons. The first was that roughly the mean and the median ages of the data was at 30. And second, because having a centered age makes interpreting $\beta_n$ easier. The value of $\beta_n$ represents how changes away from the centered age affects player $n$'s probability of hitting a HR, $\pi_{nip}$. In a normal GLM's, $\beta_n$ would not be able to adapt to each player, $n$, as it does in our model. It would only be able to consider how aging affects the average player's HR probability and then apply that effect to all players. This implicitly assumes that each player is affected identically by changes in age. In contrast, we employ multilevel modeling techniques to our model which assumes that most players experience the same general effects of aging, but allows for subtle changes depending on the player. This is the job of our hyper-priors, $\mu_1$ and $\sigma_1$. The parameter $\mu_1$ acts as the general mean which we expect most players to follow. However, since $\mu_1$ is a distribution, it allows for our model to update its posterior with observed data so that it is not heavily influenced by either a few data points or our strong priors. Our standard deviation parameter, $\sigma_1$ acts similarly. Thus, we have estimates $\beta_n$ for player $n$ which describes how aging uniquely affects their probability of hitting a HR in a given AB, $\pi_{nip}$, but these unique estimates generally follow the same trend.

>>>>

If modeling for the sake of learning was not reason enough to embark on this endeavor, there is the possibility of financial compensation also looming with this research question. If one is able to accurately predict the HR's that a player hits in a season, then they would probably be able to gain employment for a MLB team. HR's are the pinnacle of a hitter's trip to the plate, they are guaranteed to score at least one run and can lead to up to four being scored depending on how many people are on base. Thus, MLB teams which seek to maximize wins are likely looking to hire analysts who can accurately predict the HR's a player would hit so that they know who to target for drafting, signing, and trading. In addition, sports betting has recently become legal in many states across the United States. Thus, if someone is able to accurately predict HR's before a season starts, they would be able to make a large amount of money from gambling. Dr. Parson and I are more interested in the former scenario concerning gainful employment, but other readers may not share our interests and it is fair to point out the obvious consequences of this research.

@@

In Major League Baseball (MLB) the team on offense is allowed to send one player, at a time, up to home plate to "bat." Any time a player is sent up to bat this constitutes a plate appearance (PA). Depending on the outcome of the PA it may or may not be considered an at-bat (AB). If the PA ends in the player being walked (BB \[Base on Balls\]) then it is not recorded as an AB because it is thought the player did not receive a fair chance at hitting the baseball. On the other hand, if the PA ends in the player getting a hit or getting out, then it is assumed that he had a fair chance at hitting the ball and the PA is recorded as an AB. This distinction to the average fan is mostly arbitrary. However, in a modeling context it is very important. When modeling a player's ability to hit home runs (HR), as this paper is concerned with, the distinction is important because it is unfair to consider all the PA's of a player since this set of outcomes contains outcomes where it is assumed that the player has not been given a fair chance to hit the ball. For similar reasons, considering the subset of outcomes where it is assumed that they player has been given a fair chance to hit the ball seems more appropriate, this subset being AB's. Thus, this paper uses AB's as the main predictor for home runs hit by a certain player. Fellingham and Fisher (2017) also agree with this reasoning and use AB's, rather than PA's, in their research.

Furthermore, it is logical to ask: why not consider both? Incorporating more predictors can often make a model more accurate at prediction; and both Fellingham and Fisher (2017) and this paper are concerned with maximizing predictive power of our respective models. The reason this paper does not consider both is because of multicollinearity. Multicollinearity is where two independent variables in a model are highly or perfectly correlated and the model is unable to determine which variable is responsible for the corresponding change in the depending variable. Without a doubt there is a strong correlation between PA's and AB's because one is a subset of the other. In fact, we can see that the correlation between the two is quite strong in @fig-BattingCor. If we included both PA's and AB's in our model, our model's alorithm Markov Chain Monte Carlo (MCMC) would have a difficult time determining which of the two variables is responsible for the changes in the dependent variable, HR's. For other generalized linaer models (GLM's) it becomes impossible to deal with mullticolinearity and either the model parameters need to be changed or the model is deemed imossible to calculate. Thus, we neglect PA's from our model both because theory stipulates it is an inferior predictor and used in tandem with our current predictor it would cause problems for the modeling software.

An interesting quirk of baseball is that, like soccer, stadiums differ in dimensions. This is not universal; baseball fields and stadiums must have the same measurements for their infields. Yet, for the most part baseball stadiums are allowed to configure their outfields as they please. This becomes important because most HR's are achieved by hitting the baseball over the outfield fence. Thus, fences closer to home plate generally lead to more HR's for all hitters, ceteris paribus. The non-standardized nature of stadiums' fences gives them a variety of unique shapes and ranging outfield sizes. For example, Fenway Park (1923-) is known to have the closest outfield wall with their "Green Monster" in left field being only 305 ft. from home plate. While the Polo Grounds (1891-1963) boasted one of the furthest walls from home plate at 483 ft. in center field. Stadiums also differ in location. Marlins Park (2012-2020) sits at 15 ft. above sea level while Coors Field (1995-) sits at 5183 ft. above sea level. These changes in eleveation allow the ball to fly farther in the air, and hence also impact HR rates. As illustrated there are not only vastly different dimensions that baseball fields can take on but also different environments which affect a player's ability to hit home runs. Thus, any model which attempts to model players' ability to hit HR's needs to consider ballpark factors.

MLB has existed since about 1871. Since its inception the game has undergone some incredible transformations which have vastly changed the nature of the game, making it into what we know today. For example, from 1900-1920 baseball was in the "deadball era." During this period the materials used to create baseballs made it so that the ball absorbed much more force than it does today. This had a plethora of implications for the game and how it was played; for example, HR's were much less common then they are today because they were harder to hit. In contrast, the period from 1994-2004 is known informally as the "steroid era" because of the rampant use of performance enhancing drugs (PED) used by players. This, in combination with other factors such as harder materials used to make baseballs, led to players obliterating a variety of previous hitting records, such as the career and season home runs records. It is obvious from these facts that the decade played for player's had a drastic impact on their ability to produce some outcomes, such as HR's. In fact, this can be visualized by considering all the runs scored during a given season from 1871-2022 displayed in @fig-RunsByYear, or considering the total HR's hit from 1871-2022 displayed in @fig-HRByYear.

Finally, age is an important factor in athletic ability. In MLB, seniority can act both as both a blessing and a curse. On the one hand, older players are less able to hit home runs because their body is starting to lose its peak athletic performance compared to when it was, say, 30. On the other hand, these older players also have more experience than younger players and thus may be better able to hit home runs from strategies they have found during their time playing. Further we can visualize this by looking at the median HR's hit by each player from 1871-2022, which is displayed in @fig-HR.By.Age.All. However, this graph does not depict some of the nuances of the data well. For example, we have almost no data on how often 18 -year-old players hit HR's because they play in MLB infrequently. As can be seen with the standard error bars on the ages above 40, older players suffer a similar problem of small sample size. Instead, consider @fig-HR.By.Age.20to40 which only displays players who are 20-40 years old during the season. This graph clearly shows the complex non-linear relationship between player performance and age. It shows that players are generally improving in their ability to hit HR's from 20-30 years old. Then, however, we see a slight downwards trend followed by an upward trend with increasing error bars from ages 30-40. This shows that as players age past 30 their abilities start to decline. However, the jump afterwards can probably be attributed to players who are very good, continuing their careers. If a player is very good then they are going to be able to perform at above average when they are, say, 35. However, if the player is not generally considered good, then a team will likely opt to try out a new young rookie who may not be as good in the moment but is more projectable. Thus, we see less players playing past 35 which explains the higher standard errors, but they are of a higher caliber than any other age group which is why they are still playing. Thus, we see their proportion as higher than we would expect, given the trend of other 30-35 year old players.

@@

As mentioned, Fellingham and Fisher (2017) have both considered and published their thoughts concerning prediction of player home run performance using Bayesian multilevel modeling. To do this, Fellingham and Fisher (2017) used data from the Lahman data set spanning 1871 to 2008 to estimate the parametric weights for their model. Over this time frame they only considered players who had played for at least 6 seasons and who, in each of those seasons, had at least 50 AB's. They further restricted their pool of players to only considered seasons where the players were 18 years old to 45 years old.

To predict a player's HR's they used a binomial logistic model. This model predicts HR's using the AB's of a player for the number of trials and the underlying ability of a player to hit a HR, $\pi$, as their probability of success. Fellingham and Fisher (2017) estimate a player's ability to hit home runs using his age in a season, decade born, season played, and home ball park as predictors. This leaves Fellingham and Fisher (2017) with the following model, which we have adapted to make their model's parameter notation match our model's notation (in @sec-Model):

@@

The parameter's $\beta_{n,1-4}$ describe the weights of the orthogonal polynomial, up to the quartic degree, for the ages of the players. We theorize, because Fellingham and Fisher (2017) did not explicitly articulate why, they chose a quartic model because it has a closer fit to the data displayed in @fig-HR.By.Age.All and @fig-HR.By.Age.20to40. The properties of a squared term capture the rise and fall of player's HR probability over their career. The inclusion of the cubic term allows for the modeling of two peaks, while the quartic term allows it to maintain that overall trend of up and then down seen in the data. This appears in @fig-HR.By.Age.20to40 where players see a steady increase in their HR probability during their career with a peak when they are 27. Then, their probability decreases a little, followed quickly by another increase, which then leads to another peak when the player is 31. Afterwards, the probability continues in a downward trend. This will be discussed in more depth in @sec-Model but our model, which aims to improve Fellingham and Fisher's (2017) model, does not account for the cubic or quartic effects due to fear of over-fitting the age data, since the distinction between the humps is rather small.

@@

In contrast, the parameter $\xi_i$ (which is $\delta_i$ in the actual paper) represents the effect that season $i$ has on player $n$'s ability to hit a HR at stadium $p$. This $\xi$ is calculated for each season considered where $i\in\{1,…,138\}$ for the 138 seasons considered. This $\xi_i$ is calculated independent of each player so it represents the different changes that we see in HR ability over all years considered. An example of why this effect is important refer to @fig-HRByYear, which clearly shows how HR rates have differed drastically over the years considered. Although, the inclusion of $\xi_i$ is confusing because it probably captures the same effect that $\alpha_n$ captures. Most players in the MLB are within 10 years of age of each other with the most extreme different in age being about 20 years. Thus, the $\alpha_n$ of players is unlikely to differ by very much and any difference in $\alpha_n$ which appears is likely also apparent in $\xi_i$. Thus, the inclusion of both $xi_i$ and $\alpha_n$ is strange because it introduces a strong possibility of multicollinearity since both parameters describe roughly the same trend in the data.

The parameter $\delta_p$ (which is $\xi_p$ in the actual paper) is the effect that playing in park $p$ has on the HR probability $\pi_{nip}$. As discussed in @sec-BaseBack, there is strong precedent to include such a term because the variance between ballpark dimension and enviornment can be drastic. Fellingham and Fisher (2017) estimate the effect that each park has on a player's ability to hit a HR and then add the associated effect to the players who played at that ballpark in a given year.

One of the main reasons that Fellingham and Fisher's (2017) model is so strange is because of the priors that they chose. Disregarding that there is no justification for these priors in their paper, Fellingham and Fisher (2017) set the following priors for these parameters:

@@

These priors are quite strange because they are so uninformative. For example, consider the priors on either $\xi_i$ or $\delta_p$. The priors state that on a logistic scale, it is normal for the decade of birth and season of play parameters to vary along the logistic interval \[-6.5, 1.5\]. On the normal probability scale, this corresponds to probability fluctuations of about (0.0015, 0.81). Furthermore, if you only consider the means that Fellingham and Fisher have used here, then you would be left with a player who, on the logit scale, hits HR's at a probability of -7.5. On the normal probability scale this becomes about 0.00055, which is 0.055% chance any AB ends in a HR. Confusingly, this is not even close to the 2.8% that Fellingham and Fisher (2017) noted in their paper. Surely there must be better priors than this which Fellingham and Fisher (2017) could have used since it does not seem that even they believe these priors that they have set. Coincidentally, this is probably why in using MCMC to estimate their parameters they needed to use 200,000 iterations. Likely, their model could have estimated their parameters much more easily had they given it stronger priors. This is one of the areas this paper seeks to improve on from Fellingham and Fisher's (2017) methods.

Another questionable choice that Fellingham and Fisher (2017) make is that they estimated their model parameters using FORTRAN. Given the antiquity of FORTRAN, replication of Fellingham and Fisher (2017) is quite difficult. Not to mention that the 200,000 iterations makes the research so computationally expensive that it is nearly impossible for any other research team to replicate. This is disappointing because the scientific method relies on replication for the advancement of knowledge. Thus, our research seeks to make the algorithms used in this research more accessible because we chose to use R to interface with the modeling software Stan, which is written in the (more contemporary) C++ programming language. Furthermore, our model will use less iterations and stronger priors so that other research teams can more easily repicate our findings.

@@

Given all the previous problems with Fellingham and Fisher (2017) mentioned above, there is an argument to be made that a simple generalized linear model (GLM) could be used instead of their complicated Bayesian semiparametric model. While it is true that a GLM could be used to predict HR's of a player it would not nearly be as accurate as a Bayesian model. This is because the greatest strength of using Bayesian modeling is that it allows for multilevel models to be created. That is, Bayesian models allow "learning" from clusters (sometimes known as "pools") of observations when there is not as much data available on a particular obersvation. Our model employs this method much more liberally than Fellingham and Fisher (2017). In turn, this should allow for us to make stronger inferences than Fellingham and Fisher (2017) because of this "learning".

@@

In this model, we say that the $HR$'s of player $n$ follow a binomial distribution with number of trials, for player $n$ in season $i$ played at park $p$, are $AB_{nip}$. The probability of success (hitting a HR) is denoted as $\pi_{nip}$ for player $n$ in season $i$ playing at park $p$. Furthermore, we estimate $\pi$ on the logistic scale.

As discussed in @sec-BaseBack, there are many factors that influence player $n$'s underlying probability of hitting a $HR$ in a given $AB$. The first parameter which affects this probability, $\pi_{nip}$, is $\alpha_n$. In our model $\alpha_n$ can generally be thought of as the intercept term which is used to account for a player's innate ability to hit HR's. From our exploratory observations of the data, we noticed that the average (median or mean) proportion of AB's which resulted in HR's is about 0.03. This means that about 3% AB's result in HR's. This can be seen in @fig-HR.By.Age.20to40, @fig-pDist, and @fig-pDistSmall. On the logit scale, which is what our model relies on for estimation, a probability of 0.03 is about -3.5. We recognize that some players, for whatever reason, have better "innate" abilities to hit HR's than others. Therefore, we used a multilevel modeling technique for $\alpha_n$ to capture that player-to-player difference. To begin capturing this difference in $\alpha_n$ across players, we set the hyper-priors for $\alpha_n$ to be $\mu_0$ and $\sigma_0$. Where $\mu_0\sim Normal(-3.5,0.1)$ and $\sigma_0\sim Exponential(1)$. We set the mean of $\mu_0$ to be -3.5 because we estimate $\alpha_n$ on the logistic scale and we want it to be close to the average player's value. Since the parameters of $\alpha_n$ are distributions, we can think of $\alpha_n$ as a set of distributions, for the different values that either $\mu_0$ or $\sigma_0$ could take on. Further, the likelihood of the particular distribution that $\alpha_n$ takes on is depndent on the likelihood of either $\mu_0$ or $\sigma_0$ taking on a particular value. For example, $\mu_0=-3.0$ is extremely unlikely because it lies four standard deviations away from the mean of its normal distribution -3.5. Thus a distribution for $\alpha_n$ being $Normal(-3,\sigma_0)$ is highly unlikely for any value of $\sigma_0$. We chose $\mu_0$ to be normally distributed because there is precedent in Fellingham and Fisher (2017) to use normal distributions for their mean hyper-priors. Another reason it is easier to interpret normal distributions. We chose $\sigma_0$ to be exponentially distributed because the exponential distribution does not allow for negative values and the bulk of the possible values of $\sigma_0$ would be clustered near 0. Since we cannot have negative values for a standard deviation it seemed appropriate. This use of these hyper-priors is what makes the model appear to "learn". The model already comes in with certain expectations of what values $\alpha_n$ should take on, rather than simply coming into each new $n$ knowing nothing, which is what happens with more traditional GLM's. Finally, with this new set of distributions for $\alpha_n$ being the priors, our model constructs a posterior using Bayes theorem and the data on player $n$. It repeats this process for all 657 players that we considered for this research question to create a different $\alpha$ distribution for each player, $n$. Whatever value in $\alpha_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.

The next parameter we considered which affects $\pi_{nip}$ was $\beta_n$ which represents the effect centered age (the player's age minus 30) had on player $n$'s ability to hit HR's. This topic was briefly touched upon in @sec-BaseBack, where we discussed how a player's age can impact their their strategy and physical prowess. These ideas are well demonstrated in @fig-HR.By.Age.20to40. In @fig-HR.By.Age.20to40, one can easily see that, in general, players' ability to hit HR's improves over the course of their career and peaks when players reach about 30 years old. After this peak, a player's ability starts to decline for the rest of their career. We chose to center the the age of the all players at 30 for several reasons. The first was that roughly the mean and the median ages in the data was 30. Second, having a centered age makes interpreting $\beta_n$ easier. The value of $\beta_n$ represents how changes away from the centered age affects player $n$'s probability of hitting a HR, $\pi_{nip}$. In a normal GLM's, $\beta_n$ would not be able to adapt to each player, $n$, as it does in our model. It would only be able to consider how aging affects the average player's HR probability and then apply that effect to all players. This implicitly assumes that each player is affected identically by changes in age. In contrast, we employ multilevel modeling techniques to our model where $\beta_n\sim Normal(\mu_1,\sigma_1)$, $\mu_1\sim Normal(0,0.1)$, and $\sigma_1\sim Exponential(10)$. Similar to understanding $\alpha_n$, we can interpret $\beta_n$ as a collection of distributions corresponding to the different values of its hyper-priors $\mu_1$ and $\sigma_1$ and their corresponding likelihoods. We chose a normal distribution for $\mu_1$ for reasons similar to $\mu_0$. It makes interpretation easier and there is precedent in the literature. Again, for similar reasons to $\sigma_0$ we chose $\sigma_1$ to be exponentially distriubted. This ensured that $\sigma_1$ would not be negative and that the bulk of the possible values of $\sigma_1$ would be clustered near 0. With the distributions calculated for $\beta_n$ we can now use $\beta_n$ as the prior and combine it with the player data to generate the posterior using Bayes theorem. We repeat this process for all 657 players considered to get a unique distribution for each player $\beta_n$. This means that each player, $n$ has a unique aging coefficient which represents how aging effects them. Whatever value in $\beta_n$ has the maximum likelihood, that's what we used to generate our predictions for inference in @sec-Results.

Similarly $\eta_n$ is the effect that player $n$'s squared age has on his probability of hitting a HR in an AB. The purpose of this squared term is to capture the non-linear relationship between hitting a HR and player $n$'s age. It is apparent in both @fig-HR.By.Age.20to40 and @fig-HR.By.Age.All that there is a parabolic relationship between age and player $n$'s ability to hit HR's. This ability appears to rise steadily for the average player until it reaches its apex when the player is roughly 30 years old. Then, in the seasons where the player is more than 30-years-old we see a decline in HR probabilities. The shape we are left with, as can be seen in @fig-HR.By.Age.20to40, is one of an upside-down parabola. Thus, by including this squared age effect, we are able to capture that unique relationship. However, as one can easily notice in @fig-HR.By.Age.20to40, the overall shape is parabolic but for ages past 37 the data begins to trend upwards. Likely, this trend upwards is due to players who have above average skill. That is, players who have earned massive contracts because they were so much better than the average player in their prime earned large multi-million dollar contracts and teams do not want to waste the this money so they play these players. So, we are then left with this group of outlying players who do not follow the trend of data at any point over their career, but it only became noticeable after the other, more normal, players disappeared from the considered sample. Returning to the parameter $\eta_n$, it is nearly impossible to interpret what it is doing because it describes the effect of centered age squared on HR probability $\pi$, and centered age squared on its own is already almost a meaningless concept. Therefore, it is best understood that this term simply accounts for the non-linearity we see in the data and not become too concerned with understanding its estimates. However, it is easy to justify our choice for the priors and hyper-priors of $\eta_n$, $\mu_2$, and $\sigma_2$, respectively. We chose $\eta_n\sim Normal(\mu_2,\sigma_2)$ for the same reasons discussed for both $\alpha_n$ and $\beta_n$: we wanted a multilevel model which described how individual players were impacted by the parameters. We chose $\mu_2\sim Normal(0,0.01)$ because we assumed that there would be a negative effect so small that it would be very close to zero. Further, any distinction to the slightly negative of zero would be hard to justify and, in essence, arbitrary. Therefore, we went with the mean of 0 and the standard deviation of 0.01 to account for any players who have a more or less extreme non-linearity to their trajectories. That is, higher or lower peaks and troughs throughout their career for their probability $\pi$. We chose $\sigma_2\sim Exponential(100)$ because standard deviations cannot be negative, which the exponential distribution lends itself nicely to. Furthermore, the exponential distribution with a parameter of 100 is very concentrated near 0. Since we know that the coefficient of $\eta_n$ is likely to be very small since it is being multiplied by the age squared (which can be up over 100). Thus, we can be confident that our standard deviation will end up being small for our estimate. This means that the estimates our model generates will be close to 0.
